---
title: "Analyze ArrayExpress dataset"
author: "Jie Tan"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{ArrayExpress example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This is an example for analyzing a publicly available microarray dataset
from the ArrayExpress database.

# Data preparation
Load in required libraries.

```{r}
library("ADAGEpath")
library("DT")
library("dplyr")
```

Before any analysis, we need to specify the ADAGE model and the data compendium
we want to use.

```{r}
model <- eADAGEmodel
compendium <- PAcompendium
probe_dist <- probedistribution
```

Here we use the dataset
[E-GEOD-41926](https://www.ebi.ac.uk/arrayexpress/experiments/E-GEOD-41926/)
as an example. Modify the accession number to load other *P.a.* microarray
dataset from ArrayExpress.

```{r}
accession <- "E-GEOD-41926"
data_raw <- load_dataset(input = accession, isProcessed = FALSE,
                         isRNAseq = FALSE, model = model,
                         compendium = compendium,
                         quantile_ref = probe_dist,
                         download_folder = "./download", norm01 = FALSE)
```

ADAGE only accepts expression values in the (0,1) range. We linearly transform
expression values to be between 0 and 1 using the Pa compendium as the reference.

```{r}
data_normed <- zeroone_norm(input_data = data_raw, use_ref = TRUE,
                            ref_data = compendium)
```

Now we query the sample information from ArrayExpress to understand the
experimental condition of each sample.

```{r}
pheno_table <- get_sample_info(accession)
```

We reorder samples in the pheno_table to have the same order as the
expression data.

```{r}
pheno_table <- pheno_table[match(colnames(data_raw)[-1],
                                 pheno_table$`Array Data File`), ]
DT::datatable(pheno_table, class = 'cell-border stripe')
```

We set the sample phenotypes according to the genotype information.

```{r}
data_pheno <- pheno_table$`Characteristics [genotype]`
```

# ADAGE signature analysis

## Activity calculation
We calculate the activity of each signature for each sample in the dataset.

```{r}
data_activity <- calculate_activity(input_data = data_normed, model = model)
```

The returned `data_activity` is a `data.frame` with signature names in the first
column and activity values per sample starting from the second column.

## Active signature detection
We are interested in identifying differentially active signatures between
wildtype and plcH mutant.

We set indices to samples that will be included in the two-group comparison.
Here the first four samples are two wildtype and two plcH mutant. Modify the
indices to include different samples in the two-group comparison.

```{r}
indices <- 1:4
```

We use [limma](https://bioconductor.org/packages/release/bioc/html/limma.html)
to perform a differential activation test. `limma` is more robust
than a simple t test when sample size is small. A two-group limma analysis is
provided in the function `build_limma()`. You can also build other `limma` models
to test signatures' activities when the experimental design is more complex.

```{r}
limma_result <- build_limma(input_data = data_activity[, c(1, indices + 1)],
                            phenotypes = data_pheno[indices],
                            use.bonferroni = FALSE)
```

To take both absolute activity difference and significance into account, we
use pareto fronts to pick the most differentially active signatures. We extract
differentially active signatures in the first 10 layers of pareto
fronts. Modify N_fronts to get more or fewer signatures.

```{r}
active_sigs <- get_active_signatures(limma_result = limma_result,
                                     pheno_group = "both",
                                     method = "pareto", N_fronts = 10)
```

Signatures that are differentially active between wildtype and plcH mutant are:

```{r}
active_sigs
```

Check out each signature's activity changes and significance in the `limma` test.

```{r}
plot_volcano(limma_result = limma_result, highlight_signatures = active_sigs,
             interactive = TRUE)
```

Look at how the activities of active signature vary across samples.

```{r}
plot_activity_heatmap(activity = data_activity, signatures = active_sigs)
```

Based on the volcano plot and the activity heatmap, we can see that Node34pos,
Node28neg, Node41pos, Node50pos, Node69pos, Node49neg, and Node65pos are active
in wildtype, while the rest signatures become more active in plcH mutant.

## Signature overlap
To reduce the number of signatures to look at, we can check whether these active
signatures greatly overlap with others.

`plot_signature_overlap` creates a heatmap of odds ratios. The odds ratio
represents the odds that two signatures share a specific number of genes.

```{r}
signature_similarity <- plot_signature_overlap(selected_signatures = active_sigs,
                                               model = model)
```

We can see that the four signatures (Node50pos, Node28neg, Node34pos, and
Node41pos) that are most strongly active in wildtype
share significant number of genes. Node49neg overlap with Node65pos.
Node34neg and Node28pos that are active in plcH mutant also overlap.

Next we calculate the marginal activities of active signatures. Marginal
activity is defined as the activity of signature A after removing genes that
are shared with signature B.

```{r}
marginal_activity <- calculate_marginal_activity(
  input_data = data_normed,
  selected_signatures = active_sigs, model = model)
```

Again, we build a `limma` model to test whether these marginal activities
are still strongly different between two conditions.

```{r}
marginal_limma <- build_limma(input_data = marginal_activity[, c(1, indices + 1)],
                              phenotypes = data_pheno[indices])
```

Let's visualize the marginal activties in a matrix heatmap.  The value in this
matrix represents the -log10 transformed adjusted p value in
the activation test when the effect of the column signature is removed from
the row signature. Values in the diagonal of the heatmap are the activation
significance of signatures themselves. 

```{r}
plot_marginal_activation(marginal_limma_result = marginal_limma,
                         signature_order = colnames(signature_similarity),
                         sig_cutoff = 0.05)
```
Node58neg, Node116pos, Node272pos are masked by Node257pos. Node68neg is masked
by Node265neg. Node 34neg and Node28pos can mask each other. We keep Node34neg
because it has higher significance. Node50pos, Node34pos, Node41pos, and
Node49neg are masked by Node28neg.

```{r}
unique_active_sigs <- remove_redundant_signatures(marginal_limma,
                                                  sig_cutoff = 0.05)
unique_active_sigs
```

## Active signature annotation
We download *Pseudomonas aeruginosa* KEGG pathway terms from the
[TRIBE](http://tribe.greenelab.com/#/home) web server.

```{r}
KEGG <- fetch_geneset(type = "KEGG")
# we only consider KEGG pathways with more than 5 genes and less than 100 genes
# as meaningful pathways
KEGG_subset <- KEGG[lengths(KEGG) >= 5 & lengths(KEGG) <= 100]
```

We associate active signatures to known KEGG pathways.

```{r}
pathway_result <- annotate_signatures_with_genesets(
  selected_signatures = unique_active_sigs, model = model, genesets = KEGG_subset)
DT::datatable(pathway_result)
```

In this case, there are also signatures uncharacterized by KEGG.
```{r}
uncharacterized_sigs <- setdiff(unique_active_sigs, pathway_result$signature)
uncharacterized_sigs
```

Check the signature similarity of the uncharacterized signatures.
```{r}
plot_signature_overlap(selected_signatures = uncharacterized_sigs)
```
## Gene-gene network
We can check how genes in the active signatures cluster in the ADAGE gene-gene
network.

We can calculate an expression fold change for each gene and pass it to the
gene-gene network to show as node color. Again, we use `limma` to test
differential expression and get the logFC.

```{r}
data_raw_limma <- build_limma(input_data = data_raw[, c(1, indices + 1)],
                              phenotypes = data_pheno[indices],
                              use.bonferroni = FALSE)
# build a gene:fold change table from limma result
gene_logFC <- data.frame(geneID = rownames(data_raw_limma),
                         logFC = data_raw_limma$logFC)
```

Visualize the ADAGE gene-gene network of the active signatures.

```{r}
visualize_gene_network(selected_signatures = unique_active_sigs,
                       model = model, cor_cutoff = 0.5,
                       gene_color_value = gene_logFC,
                       curated_pathways = KEGG)
```

To review a signature, we can list all genes in it.

```{r}
gene_annotation <- annotate_genes_in_signatures(selected_signatures = "Node28neg",
                                                model = model,
                                                curated_pathways = KEGG)
# add expression fold change to the gene table
gene_annotation <- dplyr::right_join(gene_logFC, gene_annotation,
                                     by = c("geneID" = "LocusTag"))
DT::datatable(gene_annotation)
```

And visualize its genes with the gene-gene network.

```{r}
visualize_gene_network(selected_signatures = "Node28neg",
                       model = model, cor_cutoff = 0.5,
                       gene_color_value = gene_logFC,
                       curated_pathways = KEGG)
```

We can also review a group of similar signatures. Simply replace one signature
name with a vector of signature names.
