---
title: "Analyze a factorial design dataset"
author: "Jie Tan"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Factorial design example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This is an example for analyzing a factorial-design experiment. Factorial design
means that there are more than one variable in the experimental design and
all combinations of these variables have samples available.

# Data preparation
Load in required libraries.

```{r}
library(ADAGEpath)
library(DT)
library(limma)
```

Before any analysis, we need to specify the ADAGE model and the data compendium
we want to use.

```{r}
model <- eADAGEmodel
compendium <- PAcompendium
probe_dist <- probedistribution
```

We first load in the dataset
[E-GEOD-17296](https://www.ebi.ac.uk/arrayexpress/experiments/E-GEOD-17296/).
It has two variables in its experimental design: strain and growth phase.

```{r}
accession <- "E-GEOD-17296"
data_raw <- load_dataset(input = accession, isProcessed = FALSE,
                         isRNAseq = FALSE, model = model,
                         compendium = compendium,
                         quantile_ref = probe_dist,
                         download_folder = "./download", norm01 = FALSE)
```

ADAGE only accepts expression values in the (0,1) range. We linearly transform
expression values to be between 0 and 1 using the Pa compendium as the reference.

```{r}
data_normed <- zeroone_norm(input_data = data_raw, use_ref = TRUE,
                            ref_data = compendium)
```

To better understand the dataset, we query its sample information from
ArrayExpress. Later we define sample phenotypes based on this query result.

```{r}
pheno_table <- get_sample_info(accession)
```

We reorder samples in the pheno_table to have the same order as the
expression data.

```{r}
pheno_table <- pheno_table[match(colnames(data_raw)[-1],
                                 pheno_table$`Array Data File`), ]
DT::datatable(pheno_table, class = 'cell-border stripe')
```

# ADAGE signature analysis

## Activity calculation
We calculate the activity of each signature for each sample in the dataset.

```{r}
data_activity <- calculate_activity(input_data = data_normed, model = model)
```

The returned `data_activity` is a `data.frame` with signature names in the first
column and activity values per sample starting from the second column.

## Factorial analysis using limma
There are many statistical models for factorial design. Here we employ a
[limma](https://bioconductor.org/packages/release/bioc/html/limma.html)-based
model to help answer questions of interest. Please refer to Chapter 9.5 in limma
[usersguide](https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)
for more details.

Based on the `pheno_table`, we first create a factor specifying strains for
each sample.

```{r}
strain_pheno <- factor(c("wt", "wt", "roxSR", "roxSR", "anr", "anr",
                         "wt", "wt", "roxSR", "roxSR", "anr", "anr"))
```

We next create a factor specifying growth phase for each sample.

```{r}
phase_pheno <- factor(c(rep("exp", 6), rep("stat", 6)))
```

Collect strain/phase combinations into one factor.

```{r}
data_pheno <- factor(paste(strain_pheno, phase_pheno, sep = "."))
```

Build the design matrix.

```{r}
design <- model.matrix(~0+data_pheno)
colnames(design) <- levels(data_pheno)
```

Fit a limma model on signature activity with the design matrix.

```{r}
fit <- limma::lmFit(data_activity[, -1], design)
```

Now we need to decide the comparison we want and make contrasts. Let's say we
want to compare wildtype and anr mutant in the exponential phase; wildtype
and anr mutant in the stationary phase; and compare the difference between
the previous two comparisons.

```{r}
cont.matrix <- limma::makeContrasts(
  anrVSwtInEXP = anr.exp - wt.exp,
  anrVSwtInSTAT = anr.stat - wt.stat,
  Diff = (anr.exp - wt.exp) - (anr.stat - wt.stat),
  levels = design)
fit2 <- limma::contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)
```

Now we can check the test result of each contrast. Let's use the contrast
"anrVSwtInEXP" as an example and you can modify it to check out the result
of other contrasts.

```{r}
limma_result <- limma::topTable(fit2, coef = "anrVSwtInEXP",
                                number = nrow(data_activity), sort.by = "none")
rownames(limma_result) <- data_activity$signature
```

To take both absolute activity difference and significance into account, we
use pareto fronts to pick the most differentially active signatures. We extract
differentially active signatures in the first 5 layers of pareto
fronts. Modify N_fronts to get more or fewer signatures.

```{r}
active_sigs <- get_active_signatures(limma_result = limma_result,
                                     pheno_group = "both",
                                     method = "pareto", N_fronts = 5)
```

Signatures that are differentially active between anr mutant and wildtype
in the exponential phase are:

```{r}
print(paste(active_sigs, collapse = ","))
```

Check out each signature's activity changes and significance in the `limma` test.

```{r}
plot_volcano(limma_result = limma_result, highlight_signatures = active_sigs,
             interactive = TRUE)
```

Look at how the activities of these signature vary across samples.

```{r}
plot_activity_heatmap(activity = data_activity, signatures = active_sigs)
```

Now let's check out the contrast "Diff". It identifies the signatures that
response to anr knockout differently in exponential phase and stationary phase.

```{r}
limma_result <- limma::topTable(fit2, coef = "Diff",
                                number = nrow(data_activity), sort.by = "none")
rownames(limma_result) <- data_activity$signature
```

To take both absolute activity difference and significance into account, we
use pareto fronts to pick the most differentially active signatures. We extract
differentially active signatures in the first 5 layers of pareto
fronts. Modify N_fronts to get more or fewer signatures.

```{r}
active_sigs <- get_active_signatures(limma_result = limma_result,
                                     pheno_group = "both",
                                     method = "pareto", N_fronts = 5)
```

Signatures that response to anr knockout differently in exponential phase and
stationary phase. are:

```{r}
print(paste(active_sigs, collapse = ","))
```

Check out each signature's activity changes and significance in the `limma` test.

```{r}
plot_volcano(limma_result = limma_result, highlight_signatures = active_sigs,
             interactive = TRUE)
```

Look at how the activities of these signature vary across samples.

```{r}
plot_activity_heatmap(activity = data_activity, signatures = active_sigs)
```
