---
<<<<<<< HEAD
title: "Process RNAseq dataset"
=======
title: "Analyze processed RNAseq dataset"
>>>>>>> Add vignettes
author: "Jie Tan"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{RNAseq example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
<<<<<<< HEAD

This is an example for analyzing an RNA-seq dataset of *Pseudomonas aeruginosa*
PA14 strain. The package currently
only supports processed RNA-seq expression data: expression values of each
=======
This is an example for analyzing an RNA-seq dataset of *Pseudomonas aeruginosa*
PA14 strain. The package currently
only support processed RNA-seq expression data: expression values of each
>>>>>>> Add vignettes
sample at the gene level.

# Data preparation
Load in required libraries.
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
```{r}
library("ADAGEpath")
library("DT")
library("readr")
library("plyr")
```
<<<<<<< HEAD

Before any analysis, we need to specify the ADAGE model and the data compendium
we want to use.

=======
Before any analysis, we need to specify the ADAGE model and the data compendium
we want to use.
>>>>>>> Add vignettes
```{r}
model <- eADAGEmodel
compendium <- PAcompendium
probe_dist <- probedistribution
```
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
We use dataset
[E-GEOD-64056](https://www.ebi.ac.uk/arrayexpress/experiments/E-GEOD-64056/)
as an example here. Let's first download the zip file that stores processed
expression values for each sample and unzip it.
<<<<<<< HEAD

```{r}
download.file("https://www.ebi.ac.uk/arrayexpress/files/E-GEOD-64056/E-GEOD-64056.processed.1.zip",
              destfile = "download/E-GEOD-64056.processed.1.zip")
unzip(zipfile = "download/E-GEOD-64056.processed.1.zip",
      exdir = "download/E-GEOD-64056.processed.1")
```

This dataset contains 6 RNA-seq samples and 2 CHIP-seq samples. We next read in
all the RNA-seq samples and combine them into one data.frame.

=======
```{r}
download.file("https://www.ebi.ac.uk/arrayexpress/files/E-GEOD-64056/E-GEOD-64056.processed.1.zip", destfile = "download/E-GEOD-64056.processed.1.zip")
unzip(zipfile = "download/E-GEOD-64056.processed.1.zip",
      exdir = "download/E-GEOD-64056.processed.1")
```
This dataset contains 6 RNA-seq samples and 2 CHIP-seq samples. We next read in
all the RNA-seq samples and combine them into one data.frame.
>>>>>>> Add vignettes
```{r}
# RNAseq files end with "rpg.txt"
RNAseq_files <- list.files("download/E-GEOD-64056.processed.1/",
                             pattern = "*.rpg.txt")
RNAseq_file_paths <- file.path("download/E-GEOD-64056.processed.1/", RNAseq_files)
RNAseq_samples <- lapply(RNAseq_file_paths,
                         function(x) readr::read_tsv(x, col_names = FALSE))
RNAseq_data <-  plyr::join_all(RNAseq_samples, by = "X1")
colnames(RNAseq_data) <- c("geneID", RNAseq_files)
DT::datatable(RNAseq_data)
```
<<<<<<< HEAD

The gene IDs in this dataset are in the format of "PA14_XXXXX,symbol". We need
to clean them to only contain "PA14_XXXXX".

=======
The gene IDs in this dataset are in the format of "PA14_XXXXX,symbol". We need
to clean them to only contain "PA14_XXXXX".
>>>>>>> Add vignettes
```{r}
RNAseq_data$geneID <- sapply(RNAseq_data$geneID,
                             function(x) unlist(strsplit(x, ","))[1])
```
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
Now the `RNAseq_data` is in the right format that can be processed by the
`load_dataset` function. This processing step will take a while, because it
needs to first map PA14 gene IDs to PAO1 gene IDs. Then it needs to impute
the expression of missing genes. Finally it normalizes RNAseq expression
values to comparable ranges with microarray expression values using
[TDM](https://github.com/greenelab/TDM).
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
```{r, results = "hide"}
data_raw <- load_dataset(input = RNAseq_data, isProcessed = TRUE,
                         isRNAseq = TRUE, model = model,
                         compendium = compendium, quantile_ref = probe_dist,
                         norm01 = FALSE)
```
<<<<<<< HEAD

ADAGE only accepts expression values in the (0,1) range. We linearly transform
expression values to be between 0 and 1 using the Pa compendium as the reference.

=======
ADAGE only accepts expression values in the (0,1) range. We linearly transform
expression values to be between 0 and 1 using the Pa compendium as the reference.
>>>>>>> Add vignettes
```{r}
data_normed <- zeroone_norm(input_data = data_raw, use_ref = TRUE,
                            ref_data = compendium)
```
<<<<<<< HEAD

Now let's specify the phenotypes for each sample. It needs to be a character
vector and has the same sample order as the expression data loaded above.

```{r}
data_pheno <- c("phoB", "phoB", "tctD", "tctD", "wt", "wt")
```

=======
Now let's specify the phenotypes for each sample. It needs to be a character
vector and has the same sample order as the expression data loaded above.
```{r}
data_pheno <- c("phoB", "phoB", "tctD", "tctD", "wt", "wt")
```
>>>>>>> Add vignettes
Now this RNAseq dataset is ready for ADAGE signature analysis. Please refer
to this [vignette](User-input-example.html) on how to perform ADAGE signature
analysis.
