---
title: "Time-course example"
author: "Jie Tan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
```

Here is an example for analyzing a time-course experiment using the ADAGEpath R
package.

# Data preperation
Load in required pacakges
```{r}
library(ADAGEpath)
library(DT)
library(splines)
library(limma)
```
We first load in the dataset "E-GEOD-52445". It is a high resolution
time-serious data of Pseudomonas aeruginosa PAO1.
```{r loading in data}
# provide the accession number
accession <- "E-GEOD-52445"
# specify the ADAGE model and the data compendium to use
model <- eADAGEmodel
compendium <- PAcompendium
probe_dist <- probedistribution
# load in the dataset from ArrayExpress database
data_raw <- load_dataset(input = accession, isProcessed = FALSE,
                         isRNAseq = FALSE, model = model,
                         compendium = compendium,
                         quantile_ref = probe_dist,
                         download_folder = "./download", norm01 = FALSE)
# 0-1 normalize expression values using the Pa compendium as reference
data_normed <- zeroone_norm(input_data = data_raw, use_ref = TRUE,
                            ref_data = compendium)
```
To better understand the dataset, we query its sample information from
ArrayExpress. Later we define sample phenotypes based on this query result.
```{r specifying phenotype}
# query sample information from ArrayExpress
pheno_table <- get_sample_info(accession)
# reorder samples in the pheno_table to be the same as sample order in the
# expression data
pheno_table <- pheno_table[match(pheno_table$`Array Data File`,
                           colnames(data_raw)[-1]), ]
# print the pheno_table
DT::datatable(pheno_table[, 1:10], class = 'cell-border stripe')
```

# ADAGE signature analysis
## Signature activity calculation
We calculate the activity of each signature for each sample in the dataset.
```{r}
# calculate signature activities
data_activity <- calculate_activity(input_data = data_normed,
                                    model = model, HW_cutoff = 2.5)
# the returned data_activity is a data.frame with signature name in the first
# column and activity values starting from the second column.
```
## Large-range signatures
For datasets with complex experimental design, it is helpful to first look
at signatures that have the most dramatic activity changes across samples.
```{r}
# calculate range of each signature's activity
activity_range <- apply(data_activity[, -1], 1, function(x) diff(range(x)))
activity_range <- data.frame(signature = data_activity[, 1], activity_range,
                             stringsAsFactors = FALSE)
# first look at the top 15 signatures with largest range
large_range_sigs <- activity_range[order(activity_range$activity_range,
                                         decreasing = TRUE), "signature"][1:15]
```
Signatures that show dramatic changes across samples are:
```{r}
paste(large_range_sigs, collapse = ",")
```
View these signatures' activity changes in a heatmap.
```{r large range heatmap, fig.height=7, fig.width=7}
plot_activity_heatmap(activity = data_activity, signatures = large_range_sigs,
                      fix_color_range = TRUE)
```
As we can see in the heatmap, Node50pos, Node41pos, Node1neg, Node34pos, and
Node28neg become more active when oxygen is depleted, indicating these signatures
are related to anaerobic adaptation. On the contrary, Node31pos,
Node107pos, Node33neg, Node75pos, Node269pos become less active when oxygen is
depleted, indicating that they capture biological processes that depend on
oxygen. Node100pos, Node12neg, and Node69pos show an interesting activity
pattern. Their activities increase with time. Node5pos and Node112pos do not
show a clear pattern.

## KEGG annotation
We can associate these signatures to known KEGG pathways to better understand
them.
```{r KEGG}
KEGG <- fetch_geneset(type = "KEGG", max_size = 100, min_size = 5)
pathway_result <- annotate_signatures_with_genesets(
  selected_signatures = large_range_sigs, model = model, genesets = KEGG)
DT::datatable(pathway_result)
```
From the KEGG annotation, we can see that Node100pos, Node12neg, and Node69pos
are assocaited with sulfate-related metabolism; Node50pos, Node34pos, and
Node28neg are associated with nitrogen related metabolism; Node31pos and
Node33neg are associated with Type VI secretion system. Other signatures
are uncharacterized by KEGG.

## Signature similarity
We next check the gene overlap between these signatures.
```{r signature overlap, fig.height=7, fig.width=7}
# make a heatmap of odds ratios reflecting the gene overlaps between signatures
plot_signature_overlap(large_range_sigs)
```
Consistent with the KEGG annotation and the activity pattern, Node12neg,
Node69pos, and Node100pos have some gene overlaps; Node50pos, Node28neg,
Node34pos, and Node41pos have large gene overlaps; Node33neg, Node75pos have 
large gene overlaps; Node107pos, Node269pos, Node31pos have small gene overlaps.

## Genes in a signature
To further understand a signature of interest (e.g. Node69pos), we can list all
the genes in it. Also, you can simply replace one signature with a vector of
signatures to check out a group of signatures.
```{r genes}
gene_annotation <- annotate_genes_in_signatures("Node69pos")
DT::datatable(gene_annotation)
```

## Gene-gene network
We can also see how these genes cluster in the ADAGE gene-gene network.
```{r gene-gene network, fig.height=7, fig.width=7}
visualize_gene_network(selected_signatures = "Node69pos",
                       model = model, cor_cutoff = 0.5)
```

## Differential temporal pattern detection using limma
We can also use limma to help us detect signatures with general differences in
their temporal trends between the oxygen depletion and oxygen addition processes.
```{r limma test}
# build a numeric time phenotype vector from the pheno table
time_pheno <- as.numeric(gsub("\\D", "", pheno_table$`FactorValue [TIME]`))
# build a treatment phenotype factor
treat_pheno <- factor(pheno_table$`FactorValue [TREATMENT]`)
levels(treat_pheno) <- c("LowToHigh", "HighToLow")
# represent a time course with a cubic spline curve 
X <- ns(time_pheno, df = 3)
# build a design matrix for limma
design <- model.matrix(~treat_pheno*X)
# fit a limma model
fit <- lmFit(data_activity[, -1], design)
fit <- eBayes(fit)
# extract the fit result
limma_result <- topTable(fit, number = nrow(data_activity),
                         sort.by = "none")
limma_result$signature <- data_activity$signature
different_sigs <- limma_result[order(limma_result$F, decreasing = TRUE),
                            "signature"][1:15]

```
Signatures that show differences in temporal trends between oxygen depletion
and oxygen addition are:
```{r}
paste(different_sigs, collapse = ",")
```
View these signatures' activity changes in a heatmap.
```{r limma heatmap, fig.height=7, fig.width=7}
plot_activity_heatmap(activity = data_activity, signatures = different_sigs,
                      fix_color_range = TRUE)
```
This approach also identified Node34pos, Node28neg, Node50pos, Node41pos,
Node75pos and some other signatures that do not show large range of activity
perturbations but show significant differences in their temporal patterns
during the shift from high-to-low oxygen process and low-to-high oxygen process.
We can use the same strategies above (KEGG annotations, signature similarity
check, gene listing, gene-gene network) to invesitage these signatures.

You can also perform other statistical tests on signature activities to answer
other questions of interest.
