---
title: "Analyze a time-course dataset"
author: "Jie Tan"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Time course example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This is an example for analyzing a time-course experiment.

# Data preparation
Load in required libraries.
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
```{r}
library(ADAGEpath)
library(DT)
library(splines)
library(limma)
```
<<<<<<< HEAD

Before any analysis, we need to specify the ADAGE model and the data compendium
we want to use.

=======
Before any analysis, we need to specify the ADAGE model and the data compendium
we want to use.
>>>>>>> Add vignettes
```{r}
model <- eADAGEmodel
compendium <- PAcompendium
probe_dist <- probedistribution
```
<<<<<<< HEAD

We first load in the dataset
[E-GEOD-52445](https://www.ebi.ac.uk/arrayexpress/experiments/E-GEOD-52445/).
It is a high resolution time-series data of *Pseudomonas aeruginosa* PAO1.

=======
We first load in the dataset
[E-GEOD-52445](https://www.ebi.ac.uk/arrayexpress/experiments/E-GEOD-52445/).
It is a high resolution time-serious data of *Pseudomonas aeruginosa* PAO1.
>>>>>>> Add vignettes
```{r}
accession <- "E-GEOD-52445"
data_raw <- load_dataset(input = accession, isProcessed = FALSE,
                         isRNAseq = FALSE, model = model,
                         compendium = compendium,
                         quantile_ref = probe_dist,
                         download_folder = "./download", norm01 = FALSE)
```
<<<<<<< HEAD

ADAGE only accepts expression values in the (0,1) range. We linearly transform
expression values to be between 0 and 1 using the Pa compendium as the reference.

=======
ADAGE only accepts expression values in the (0,1) range. We linearly transform
expression values to be between 0 and 1 using the Pa compendium as the reference.
>>>>>>> Add vignettes
```{r}
data_normed <- zeroone_norm(input_data = data_raw, use_ref = TRUE,
                            ref_data = compendium)
```
<<<<<<< HEAD

To better understand the dataset, we query its sample information from
ArrayExpress. Later we define sample phenotypes based on this query result.

```{r}
pheno_table <- get_sample_info(accession)
```

We reorder samples in the pheno_table to have the same order as the
expression data.

=======
To better understand the dataset, we query its sample information from
ArrayExpress. Later we define sample phenotypes based on this query result.
```{r}
pheno_table <- get_sample_info(accession)
```
We reorder samples in the pheno_table to have the same order as the
expression data.
>>>>>>> Add vignettes
```{r}
pheno_table <- pheno_table[match(colnames(data_raw)[-1],
                                 pheno_table$`Array Data File`), ]
DT::datatable(pheno_table[, 1:10], class = 'cell-border stripe')
```

# ADAGE signature analysis

## Activity calculation
We calculate the activity of each signature for each sample in the dataset.
<<<<<<< HEAD

```{r}
data_activity <- calculate_activity(input_data = data_normed, model = model)
```

The returned `data_activity` is a `data.frame` with signature names in the first
=======
```{r}
data_activity <- calculate_activity(input_data = data_normed, model = model)
```
The returned `data_activity` is a `data.frame` with signature name in the first
>>>>>>> Add vignettes
column and activity values per sample starting from the second column.

## Most responsive signatures
For datasets with complex experimental design, it is helpful to first look
at signatures that have the most dramatic activity changes across samples.

We first calculate the range of each signature's activity.
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
```{r}
activity_range <- apply(data_activity[, -1], 1, function(x) diff(range(x)))
activity_range <- data.frame(signature = data_activity[, 1], activity_range,
                             stringsAsFactors = FALSE)
```
<<<<<<< HEAD

Get the top 15 signatures with largest ranges.

=======
Get the top 15 signatures with largest ranges.
>>>>>>> Add vignettes
```{r}
large_range_sigs <- activity_range[order(activity_range$activity_range,
                                         decreasing = TRUE), "signature"][1:15]
```
<<<<<<< HEAD

Signatures that show dramatic changes across samples are:

```{r}
paste(large_range_sigs, collapse = ",")
```

View these signatures' activity changes in a heatmap.

```{r}
plot_activity_heatmap(activity = data_activity, signatures = large_range_sigs)
```

=======
Signatures that show dramatic changes across samples are:
```{r}
paste(large_range_sigs, collapse = ",")
```
View these signatures' activity changes in a heatmap.
```{r}
plot_activity_heatmap(activity = data_activity, signatures = large_range_sigs)
```
>>>>>>> Add vignettes
As we can see in the heatmap, Node50pos, Node41pos, Node1neg, Node34pos, and
Node28neg become more active when oxygen is depleted, indicating these signatures
are related to anaerobic adaptation. On the other side, Node31pos,
Node107pos, Node33neg, Node75pos, Node269pos become less active when oxygen is
depleted, indicating that they capture biological processes that depend on
oxygen. Node100pos, Node12neg, and Node69pos show an interesting activity
pattern. Their activities increase with time. Node5pos and Node112pos do not
show a clear pattern.

## KEGG annotation
We can associate these signatures to known KEGG pathways to better understand
them.
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
```{r}
KEGG <- fetch_geneset(type = "KEGG", max_size = 100, min_size = 5)
pathway_result <- annotate_signatures_with_genesets(
  selected_signatures = large_range_sigs, model = model, genesets = KEGG)
DT::datatable(pathway_result)
```
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
From the KEGG annotation, we can see that Node100pos, Node12neg, and Node69pos
are associated with sulfate-related metabolism; Node50pos, Node34pos, and
Node28neg are associated with nitrogen related metabolism; Node31pos and
Node33neg are associated with Type VI secretion system. Other signatures
are uncharacterized by KEGG.

## Signature overlap
We next check the gene overlap between these signatures.

`plot_signature_overlap` creates a heatmap of odds ratios. The odds ratio
represents the odds that two signatures share a specific number of genes.
<<<<<<< HEAD

```{r}
plot_signature_overlap(large_range_sigs)
```

=======
```{r}
plot_signature_overlap(large_range_sigs)
```
>>>>>>> Add vignettes
Consistent with the KEGG annotation and the activity pattern, Node12neg,
Node69pos, and Node100pos have some gene overlaps; Node50pos, Node28neg,
Node34pos, and Node41pos have large gene overlaps; Node33neg, Node75pos have
large gene overlaps; Node107pos, Node269pos, Node31pos have small gene overlaps.

Let's take the "Node12neg, Node69pos, and Node100pos" signature cluster as
an example to calculate their marginal activities.
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
```{r}
similar_sigs <- c("Node12neg", "Node69pos", "Node100pos")
marginal_activity <- calculate_marginal_activity(input_data = data_normed,
                                                 selected_signatures = similar_sigs,
                                                 model = model)
```

View these signatures' marginal activity changes in a heatmap.
<<<<<<< HEAD

```{r}
plot_activity_heatmap(activity = marginal_activity)
```

=======
```{r}
plot_activity_heatmap(activity = marginal_activity)
```
>>>>>>> Add vignettes
From the heatmap we see that Node100pos and Node69pos still show a temporal
pattern even after genes shared with each other or Node12neg have been removed.
The temporal pattern of Node12neg becomes much weaker when Node100pos or Node69pos
is removed. Therefore, among these three signatures we can safely drop Node12neg.

## Gene-gene network
We can check how genes in these signatures cluster in the ADAGE gene-gene
network.
<<<<<<< HEAD

=======
>>>>>>> Add vignettes
```{r}
visualize_gene_network(selected_signatures = large_range_sigs,
                       model = model, cor_cutoff = 0.5)
```

## Differential temporal pattern detection using limma
We can also use customized statistical tests on signature activities to identify
signatures of interest. We apply the limma model for time-course experimental
design here as an example (Chapter 9.6.2 in limma
[usersguide](https://bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)).
This test helps us detect signatures with general differences in
their temporal trends between the oxygen depletion and oxygen addition processes.

We first create a numeric vector representing time.
<<<<<<< HEAD

```{r}
time_pheno <- as.numeric(gsub("\\D", "", pheno_table$`FactorValue [TIME]`))
```

We next create a treatment factor specifying the oxygen depletion vs oxygen
addition treatment.

=======
```{r}
time_pheno <- as.numeric(gsub("\\D", "", pheno_table$`FactorValue [TIME]`))
```
We next create a treatment factor specifying the oxygen depletion vs oxygen
addition treatment.
>>>>>>> Add vignettes
```{r}
treat_pheno <- factor(pheno_table$`FactorValue [TREATMENT]`)
levels(treat_pheno) <- c("LowToHigh", "HighToLow")
```
<<<<<<< HEAD

Represent a time course with a cubic spline curve.

```{r}
X <- ns(time_pheno, df = 3)
```

Build a design matrix for limma and then fit a limma model.

=======
Represent a time course with a cubic spline curve.
```{r}
X <- ns(time_pheno, df = 3)
```
Build a design matrix for limma and then fit a limma model.
>>>>>>> Add vignettes
```{r}
design <- model.matrix(~treat_pheno*X)
fit <- lmFit(data_activity[, -1], design)
fit <- eBayes(fit)
```
<<<<<<< HEAD

Extract the top 15 most significant signatures from the limma fit result.

=======
Extract the top 15 most significant signatures from the limma fit result.
>>>>>>> Add vignettes
```{r}
limma_result <- topTable(fit, number = nrow(data_activity),
                         sort.by = "none")
limma_result$signature <- data_activity$signature
different_sigs <- limma_result[order(limma_result$F, decreasing = TRUE),
                               "signature"][1:15]
```
<<<<<<< HEAD

Signatures that show differences in temporal trends between oxygen depletion
and oxygen addition are:

```{r}
paste(different_sigs, collapse = ",")
```

View these signatures' activity changes in a heatmap.

```{r}
plot_activity_heatmap(activity = data_activity, signatures = different_sigs)
```

This approach also identified Node34pos, Node28neg, Node50pos, Node41pos,
Node75pos and some other signatures that do not show a large range of activity
=======
Signatures that show differences in temporal trends between oxygen depletion
and oxygen addition are:
```{r}
paste(different_sigs, collapse = ",")
```
View these signatures' activity changes in a heatmap.
```{r}
plot_activity_heatmap(activity = data_activity, signatures = different_sigs)
```
This approach also identified Node34pos, Node28neg, Node50pos, Node41pos,
Node75pos and some other signatures that do not show large range of activity
>>>>>>> Add vignettes
perturbations but show significant differences in their temporal patterns
during the shift from high-to-low oxygen process and low-to-high oxygen process.
We can use the same strategies above (KEGG annotations, signature overlap
check, gene-gene network) to investigate these signatures.
